---
title: "Day 8: SQL"
subtitle: "IMSI Data Science Summer Workshop Series 2025"
author: "Awoyemi & Babyale"
date: today
format: 
  html:
    embed-resources: true
    toc: true
editor: 
  markdown: 
    wrap: 72
---

# Exercise 1: Places - Designing the Place of the Future

Note, from Exercise 1 to Exercise 3, we make the assumption that
`db_places` is the name of the data base connection. Recall that Quarto
supports the use of a SQL engine so long as a connection is specified in
the code chunk options.

In particular, please make sure code chunks are constructed using

```{{sql}}
--| label: my-label
--| connection: db_places

-- SQL Code
```

------------------------------------------------------------------------

In this exercise, we explore setting up a database and creating a table.

-   Establish a connection to the file-backed database `my_data.sqlite`.
-   Design a table called `MyPlaces` with the following fields:
    -   `Name`: a varchar field containing the name of the place.
    -   `Category`: a varchar field that describes the place based on
        the top categories from
        [Yelp](https://blog.yelp.com/businesses/yelp_category_list/).
    -   `TravelTime`: an integer from the Chicago to the place.
    -   `DateVisited`: a varchar field that is interpreted as a date
        (**YYYY-MM-DD**) containing the last time you visited the place.
    -   `Recommendation`: an integer field that is interpreted as a
        boolean containing whether you would recommend the place (1) or
        not (0).

::: callout-note
If you are on Windows, please use the in-memory option:
:::

```{r}
library(DBI)

if(file.exists("MyPlaces.sqlite")) delete_old <- file.remove("MyPlaces.sqlite")

db_places = DBI::dbConnect(RSQLite::SQLite(), name = ":memory:")

```

```{sql}
--| label: create-table
--| connection: db_places


CREATE TABLE MyPlaces (
Name VARCHAR,
Category VARCHAR,
TravelTime INTEGER,
DateVisited VARCHAR,
Recommendation INTEGER,
PRIMARY KEY (Name)
);

```

# Exercise 2: Places - Populating Places

Within this exercise, the focus is on populating or filling the table
just designed.

-   Populate the tables created in **Exercise 1** with at least 7
    places.
-   Please make sure there is at least one place for each of the
    following categories:
    -   recommended, not recommended, and has yet to receive a
        recommendation (`NULL`),
    -   within 10 minutes, 45 minutes, and 180 minutes; and,
    -   visited within the week, the month, and the year.

```{sql}
--| label: insert-into-table
--| connection: db_places

INSERT INTO MyPlaces VALUES
("Lubbock", "Education", 10, "2025-06-10", 1),
("Amarillo", "Automotive", 180, "2025-01-15", 0),
("Ransom Canyon", "Education", 10, "2025-05-24", 0),
("Little Field", "Food", 45, "2025-06-07", NULL),
("Wolfforth", "Entertainment", 45, "2025-03-21", 1),
("Boise", "Active Life", 180, "2025-04-20", 1),
("Garden City", "Entertainment", 45, "2025-05-31", NULL);

```

```{r}

dbReadTable(db_places, "Myplaces")
```

# Exercise 3: Places - Querying Places

In the last exercise in this series, we seek to query the table and
understand what data is present.

-   Write multiple queries to learn about the data in the `MyPlace`
    table that answer the following questions:
    1.  What places require less than 45 minutes of travel time from
        Chicago?
    2.  What places would you not recommend?
    3.  How frequently does each category appear in the data?
    4.  What places have you not visited in the last week?
-   Write a single query that returns a count of the categories of
    places that you would recommend.
-   Write a single query that identifies places you recommend but have
    not visited for at least one month.
-   Write a query that lists the name of the place and the
    recommendation text that matches:
    -   If you would recommend the place, the text should read: "Seal of
        Approval by <Firstname>!"
    -   If you would **not** recommend the place, the text should read:
        "Not the best place to visit..."

```{sql connection = db_places}
--| label: Places-Querying-Places1
--| connection: db_places

SELECT Name, TravelTime FROM MyPlaces WHERE TravelTime < 45;

```

```{sql connection = db_places}
--| label: Places-Querying-Places2
--| connection: db_places

SELECT Name, Recommendation FROM MyPlaces WHERE Recommendation == 0;

```

```{sql connection = db_places}
--| label: Places-Querying-Places3
--| connection: db_places

SELECT Category, COUNT(Category) as Frequency FROM MyPlaces GROUP BY Category;

```

```{sql connection = db_places}
--| label: Places-Querying-Places4
--| connection: db_places

SELECT Name, DateVisited FROM MyPlaces WHERE JULIANDAY("2025-06-13") - JULIANDAY(DateVisited) > 7;
```

```{sql connection = db_places}
--| label: Places-Querying-Places5
--| connection: db_places

-- SELECT Recommendation, Category FROM MyPlaces WHERE Recommendation == 1;

SELECT Recommendation, COUNT(Category) AS Count FROM MyPlaces WHERE Recommendation == 1;

```

```{sql connection = db_places}
--| label: Places-Querying-Places6
--| connection: db_places

SELECT * FROM MyPlaces WHERE Recommendation == 1 and JULIANDAY("2025-06-13") - JULIANDAY(DateVisited) > 30;
```

```{sql connection = db_places}
--| label: Places-Querying-Places7
--| connection: db_places

SELECT Name,
CASE
    WHEN Recommendation = 1 THEN 'Seal of Approval by !'
    ELSE 'Not the best place to visit…'
END AS RecommendationText
FROM MyPlaces;
```

```{sql connection = db_places}
--| label: Places-Querying-Places8
--| connection: db_places

SELECT Name, Recommendation FROM MyPlaces WHERE Recommendation == 0;
```

# Exercise 4: Batter - DBI

Note, from Exercise 4 to Exercise 6, we make the assumption that
`db_lahman` is the name of the data base connection. Recall that Quarto
supports the use of a SQL engine so long as a connection is specified in
the code chunk options.

In particular, please make sure code chunks are constructed using

```{{sql}}
--| label: my-label
--| connection: db_lahman

-- SQL Code
```

## (a) Obtaining Data

Download the Lahman baseball data set from:
<https://github.com/coatless/raw-data/blob/main/lahman2016.sqlite?raw=true>
using *R*'s `download.file()` function.

Add the `cache: TRUE` code chunk option to avoid redownloading the file
multiple times when rendering the Quarto document

```{r}
#| label: download-lahman
#| cache: true

# Code here
```

## (b) Connecting to the Database

Create a connection to the `lahman2016.sqlite` SQLite database called
`db`.

```{r}
# Code here
```

## (c) Determining Tables

List all the tables that are inside the `db_lahman` connection.

```{r}
# Code here

```

## (d) Naming Fields

What are the names of all the columns in the `Salaries` and `Teams`
tables?

```{r}
# Code here
```

## (e) Pulling data into R

Retrieve the `salaries` table using the `db_lahman` connection and bring
it into *R* using the same name.

```{r}
# code here
```

# Exercise 5: Batter - Querying

## (a)

Create a SQL query embedded in Quarto that counts the average number of
home runs in the entire `Batting` table.

*Hints*:

-   You will need to use: `SELECT`, `AVG`, and `AS`.
-   The result should only contain **one** field -- `average_hr` -- and
    **1** record.

## (b)

Improve the SQL query developed in **b** and embedded in the Quarto
document by:

-   calculating the average for **each team**,
-   show only teams that have an average equal to or greater than
    **3.5**, and
-   listing the name of the team.

*Hints*:

-   You will need to use: `SELECT`, `AVG`, `AS`, `JOIN`, `GROUP BY`, and
    `HAVING`.
-   The result should only contain **two** fields -- `name` and
    `average_hr` -- and **16** records.
-   The `name` column in `Teams` contains the team name.
    -   Join the `Teams` and `Batting` tables by `teamID` and `yearID`.
-   To handle the `ambiguous column name` error requires specifying the
    table name prior to using the variable in the `SELECT ...` portion,
    e.g. `table.variable`

## (c)

Modify the **c** SQL query embedded in Quarto so that it tabulates the
average number of home runs for each player in the `Batting` table:

-   **from 1975 onwards**,
-   **takes only players with an overall average number of homeruns to
    be greater than 10**, and
-   **sorts them in descending order**.

*Hints*:

-   You will need to use: `SELECT`, `AVG()`, `AS`, `WHERE`, `GROUP BY`,
    `HAVING`, `ORDER BY`, and `DESC`.
-   The result should only contain **two** fields -- `playerID` and
    `average_hr` -- and **548** records.
-   Recall the difference between `HAVING` and `WHERE`.

# Exercise 6: Batter - Compute and Retrieve

## (a)

Write a SQL query embedded in Quarto that computes the total salary
expenditures for each team for the year 2016.

*Hints*:

-   You will need to use: `SELECT`, `SUM`, `AS`, `WHERE`, and
    `GROUP BY`.
-   The result should only contain **two** fields -- `teamID` and
    `total_salary` -- and **30** records.
-   Set the `output.var` code chunk option to be equal to
    `salary_teamid_info` so that the `data.frame` is available to *R*.

## (b)

Graph the results from **a** on a bar plot created using
[`geom_col()`](https://ggplot2.tidyverse.org/reference/geom_bar.html)
from [`ggplot2`](https://ggplot2.tidyverse.org/). Consider looking at
examples on [R Graph Gallery](https://www.r-graph-gallery.com/) for help
in constructing the plot.

By looking at the graph, what team had the highest payouts to their
players in 2016?

*Hint*: Rotate text on the x-axis of ggplot with
`theme(axis.text.x = element_text(angle = 45, hjust = 1))`.

## (c)

Instead of calculating only a single year, compute the total salaries
for each year spanning from 1990 to 2016. Graph the result on a line
plot using
[`geom_line()`](https://ggplot2.tidyverse.org/reference/geom_line.html)
from [`ggplot2`](https://ggplot2.tidyverse.org/). Graph each team's line
in a different column.

*Hints*:

-   You will need to use: `SELECT`, `SUM`, `AS`, `WHERE`, and
    `GROUP BY`.
    -   Psst, to specify multiple groups place a comma between them!
-   The result should only contain **three** fields -- `teamID`,
    `yearID`, and `total_salary` -- and **788** records.
-   Set the `output.var` code chunk option to be equal to
    `salary_teamid_yearly` so that the `data.frame` is available to *R*.
